Sparse Spectral Methods and P-Finite Element Methods on Disk-Splices and Rhombi - Part I: Zero Boundary Conditions

### 2018-12-14 papers/halfdisk/halfdisk.tex
Further edits. Added a couple of plots.

@@

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\) and \(T^{(0,1)\to(1,1)}\) such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y)
\end{align}

We can then obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}
The sparsity of the these operators can be shown using a simply integration by parts argument, again showing that the sparsity for differential operators is not dependent on the domain.

@@

@@

>>>>

\graphicspath{ {images/} }

@@

\begin{figure} 
\center
\begin{tikzpicture} 
% Box/square (4x4)
\draw[black,solid,ultra thin] (0,0)--(0,4);
\draw[black,solid,ultra thin] (0,0)--(4,0);
\draw[black,solid,ultra thin] (4,0)--(4,4);
\draw[black,solid,ultra thin] (0,4)--(4,4);
% Dots
\draw[black,fill=black] (0,0) circle (.5ex);
\draw[black,fill=black] (2,2) circle (.5ex);
\draw[black,fill=black] (4,4) circle (.5ex);
\draw[black,fill=black] (0,2) circle (.5ex);
\draw[black,fill=black] (0,4) circle (.5ex);
% Arrows
\draw[black,thick,->] (0.2,0.2)--(1.8,1.8);
\draw[black,thick,->] (2.2,2.2)--(3.8,3.8);
\draw[black,thick,->] (0,0.2)--(0,1.8);
\draw[black,thick,->] (0,2.2)--(0,3.8);
\draw [black,thick,->] (0.2,4.2) to [out=20,in=160] (3.8,4.2);
% Node (parameter) labels
\draw[] (0,0) node[anchor=east] {$(0,0)$};
\draw[] (0,4) node[anchor=east] {$(0,2)$};
\draw[] (4,0) node[anchor=west] {$(2,0)$};
\draw[] (4,4) node[anchor=west] {$(2,2)$};
% Arrow labels
\draw[] (1,1) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (3,3) node[anchor=north] {$\tfrac{\partial}{\partial x}$};
\draw[] (0,1) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
\draw[] (0,3) node[anchor=west] {$\tfrac{\partial}{\partial y}$};
% The key and headings
\draw[black,thick,->] (-2.5,3.4)--(-2.5,4);
\draw[black,thick,->] (-2.5,3.4)--(-1.9,3.4);
\draw[] (-2.3,3.2) node[anchor=west] {$a$};
\draw[] (-2.7,3.6) node[anchor=south] {$b$};
\draw[] (2,4.6) node[anchor=south] {basis conversion};
\end{tikzpicture} 
\caption{The Laplace operator acting on vectors of $P_{n,k}=P_{n,k}^{(0,0)}$ coefficients has a sparse matrix representation if the range is represented as vectors of $\smash{P^{(2,2)}_{n,k}}$ coefficients. Here, the arrows indicate that the corresponding operation has a sparse matrix representation when the domain is $\smash{P_{n,k}^{(a,b)}}$ coefficients, where $(a,b)$ is at the tail of the arrow, and the range is $\smash{P_{n,k}^{(\tilde{a},\tilde{b})}}$ coefficients, where $(\tilde{a},\tilde{b})$ is at the head of the arrow. %For example, $\smash{\frac{\partial}{\partial x}}$ has a sparse matrix representation when the domain is represented in $\smash{P_{n,k}^{(0,0,0)}}$ (resp.~$\smash{P_{n,k}^{(1,0,1)}}$) and the range in $\smash{P_{n,k}^{(1,0,1)}}$ (resp.~$\smash{P_{n,k}^{(2,0,2)}}$).
}
\label{fig:Laplace} 
\end{figure}

For general \(a,b\) and general function \(f(x,y) = \mathbf{f}^T \: \bigPab(x,y)\), there exist Banded-Block-Banded matrix operators \(D_x^{(a,b)}, \: D_y^{(a,b)}, \: W_x^{(a,b)}, \: W_y^{(a,b)}\) such that:
\begin{align}
\dx[f(x,y)] &= \bigP^{(a+1,b+1)}(x,y)^T \: D_x^{(a,b)} \: \mathbf{f}, \\
\dy[f(x,y)] &= \bigP^{(a,b+1)}(x,y)^T \: D_y^{(a,b)} \: \mathbf{f}, \\
\dx[\Wab(x,y) \: f(x,y)] &= W^{(a-1,b-1)}(x,y) \: \bigP^{(a-1,b-1)}(x,y)^T \: W_x^{(a,b)} \: \mathbf{f}, \\
\dy[\Wab(x,y) \: f(x,y)] &= W^{(a,b-1)}(x,y) \: \bigP^{(a,b-1)}(x,y)^T \: W_y^{(a,b)} \: \mathbf{f}.
\end{align}
The incrementing and decrementing of parameters as seen here is analogous to other well known orthogonal polynomials families' derivatives, for example the Jacobi polynomials, as seen in the DLMF \cite{DLMFDerivatives}.
Further, we note that there exist Banded-Block-Banded matrix operators \(T^{(1,0)\to(0,0)}\), \(T^{(0,1)\to(1,1)}\) and $T^{(0,2)\to(2,2)}$ such that:
\begin{align}
x \bigP^{(1,0)}(x,y) &= \Big(T^{(1,0)\to(0,0)} \Big)^T \: \bigP^{(0,0)}(x,y) \\
\bigP^{(0,1)}(x,y) &= \Big(T^{(0,1)\to(1,1)} \Big)^T \: \bigP^{(1,1)}(x,y) \\
\bigP^{(0,2)}(x,y) &= \Big(T^{(0,2)\to(2,2)} \Big)^T \: \bigP^{(2,2)}(x,y)
\end{align}

We can then obtain the matrix operator for \(\Delta\), that will take us from coefficients for expansion in the weighted space
\[
\Wii(x,y) \: \bigPii(x,y)
\]
to coefficients in the non-weighted space
\[
\bigPii(x,y).
\]
Importantly, this operator will have Banded-Block-Banded structure, and hence will be sparse. Note that this construction will ensure the imposition of the Dirichlet zero boundary conditions on $\Omega$. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    \Delta := D_x^{(0,0)} \: W_x^{(1,1)} + T^{(0,1)\to(1,1)} \: D_y^{(0,0)} \: T^{(1,0)\to(0,0)} \: W_y^{(1,1)}.
\end{align}

Further, we can also obtain the matrix operator for \(\Delta\) that will take us from coefficients for expansion in the space
\[
\bigPoo(x,y)
\]
to coefficients in the space
\[
\bigP^{(2,2)}(x,y).
\]
Importantly, again, this operator will have Banded-Block-Banded structure, and hence will be sparse. The matrix operator \(\Delta\) acting on the coefficients vector is then given by
\begin{align}
    D := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(0,2)\to(2,2)} \: D_y^{(0,1)} \: D_y^{(0,0)}.
\end{align}

The sparsity of the these operators can be shown using a simply integration by parts argument, again showing that the sparsity for differential operators is not dependent on the domain.

@@

\begin{figure}
	\includegraphics[scale=0.4]{example4coeffs}
\centering
\caption{Plot of the difference between the previous solution's coefficients and the current solution's coefficients under increasing degree of approximation (number of coefficients used) for $N = 2,\dots,10$, using RHS of $f(x,y) = 2(1-x^2-y^2)\cos(x) - 8x\sin(x) - x(1-x^2-y^2)\sin(x) - 4x^2 \cos(x) - 12xy$ for the Poisson equation. The solution is converging as seen by the decreasing (absolute value) difference of coefficient values.}
\centering
\label{fig:poisson}
\end{figure}

@@

\begin{figure}
	\includegraphics[scale=0.4]{example2timesteppingcovnvergence}
\centering
\caption{Plot of the current iteration's coefficients for the heat equation example, using degree of approximation $N = 5$ with initial condition of $u_0(x,y) = \Wii(x,y)$ and step size $h=0.01$, plotted for iterations $t=100, 120, \dots, 200$. The solution tends to zero, as seen by the decreasing (absolute value) coefficient values.}
\centering
\label{fig:heateqn}
\end{figure}
